---
title: "Larval Growth Heritability Estimation"
output:
  github_document:
    pandoc_args: --webtex
    toc: true
    toc_depth: 2
---

# Overview

Summary of the model outputs for the **MCMCglmm** models used for estimating heritability of larvae growth. Growth is represented here as larval area. Below I explore several different models in order to evaluate the impact of different random effects, priors, and data on heritability estimates.

In this summary the full model looks like this;

$$Area = G + M + J + S + R$$
where $G$ is pedigree or genetic component, $R$ is the damID or maternal component, $J$ is the JarID, $S$ is the seatableID, and $R$ is the residual variation. Additional simpler models are also explored (details under *Summary of Models*).

**Note** : Area is the response variable used in these models for the moment to represent larvae growth, until I confirm that the growthrate measurements in the dataset aren't corrected by egg size. Area was also scaled using the ```scale()``` function.

## Data and Scripts
```{r}
mod_output <- readRDS("~/Desktop/GrowthHeritabilityEstimate.RDS")
```

**Links to Data and Scripts**

* [Data Format Script](https://github.com/epigeneticstoocean/2018OAExp_larvae/blob/master/src/dataFormatScript.R)
* **Formatted Data** ([Individual](https://github.com/epigeneticstoocean/2018OAExp_larvae/blob/master/input_files/IndHeritabilityData.RDS),[Family](https://github.com/epigeneticstoocean/2018OAExp_larvae/blob/master/input_files/JarHeritabilityData.RDS))
* [Model Script](https://github.com/epigeneticstoocean/2018OAExp_larvae/blob/master/src/HeritabilityEstimation/L18_heritabilityOfGrowth_MCMCGlMM_Feb2021.R)

## Libraries
```{r}
library(ggplot2,quietly = T)
library(cowplot,quietly = T)
library(knitr,quietly = T)
library(MCMCglmm,quietly = T)
library(reshape2,quietly = T)
```

# Summary of Models
 
| Model | Fixed Effects | Random Effects | Description |
|:------|:--------------|:---------------|:------------|
| Model 1 | ~ 1 | ~ animal, dam, sire, Jar, and Seatable | Full Model (including sire which is not typically done) |
| Model 2 | ~ 1 | ~ animal | Simply model | 
| Model 3 | ~ 1 | ~ animal, dam, and Jar | Single experimental effect, jar |
| Model 4 | ~ 1 | ~ animal, dam, and Seatable | Single experiment effect, Seatable |
| Model 5 | ~ 1 | ~ animal and dam | Typical animal model with dam (no experimental effect) |
| Model 6 | ~ 1 | ~ animal, dam, jar, and Seatable | Full model alternative (no sire) |

# Parameterization

All models used the same set of parameters. Prior and random effects varied for each model.

* **Iterations** - 1,000,000
* **Burn-in** - 10,000
* **Thinning parameter** - 1,000

## Priors

The same prior was applied to each random effect, the specific type of prior depends on the model.

* **Cauchy Prior** - Used in some heritability models. Favored when estimated variance of the components is small.
  * Parameters (Same parameters used by Hadley et al. 2010 for this prior)
    * V : 1
    * nu : 1
    * alpha.mu : 0
    * alpha.V : 1000
* **Inverse-Wilshart** - Standard uninformative prior used in many bayesian models. May misbehave when variance is small.
  * Parameters
    * V : 1
    * nu : 0.002 (very small belief in prior means it should not inform posterior distribution)

# Complete Summary Output Table

**Note** : Model 4 and 5 were run twice for each data subset. The second use was performed using the family level summarized data (growth averaged by jar). This was done to examine the effect of estimating heritability on family level rather than individual level estimates. This approach was recent used by Tasoff et al. (2018) to look a family level heritability in Grunion in ambient and OA conditions.

```{r echo=F}
sumTab <- NULL 
## Diagnostics
for(i in mod_output){
  for(j in i){
    sumTab <- rbind(sumTab,j$summary)
  }
}
sumTab$model[sumTab$model == "Model3"] <- rep(c("Model3","Model3_InvW"),time=4)
sumTab$model[sumTab$model == "Model4"] <- rep(c("Model4","Model4_fam"),time=4)
sumTab$model[sumTab$model == "Model5"] <- rep(c("Model5","Model5_fam"),time=4)
kable(sumTab)
```

# Across Model Heritability Comparisons

```{r echo=F,fig.width=12}
sumTab$Parent_Env_Descriptive <- paste0("Parent_",sumTab$Parent_Env)
sumTab$Offspring_Env_Descriptive <- paste0("Offspring_",sumTab$Offspring_Env)
ggplot(sumTab,aes(model,animal_H2_mean,colour=model)) +
  ylim(0,1) +
  labs(y="Narrow-sense heritability",x="") +
  facet_grid(rows=vars(Parent_Env_Descriptive),cols=vars(Offspring_Env_Descriptive)) +
  geom_point() + 
  geom_errorbar(aes(ymin=animal_H2_lower, ymax=animal_H2_upper)) +
  theme_cowplot() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1))
```

### Main Takeaways

1) **Model 2 vs other models** - Not including dam (and to lesser extend one or both of the experimental random effects) has a substantial impact on the genetic component in the model (i.e. dam and other random effects consume a substantial amount of the variance explained by the genetic component when included in the model).
2) **Model 3 vs Model 3 InvW** - The two priors do not seem to alter our estimate of heritability.
3) **Model 4 vs Model 4 fam & Model 5 vs Model 5 fam** - Measuring heritability at the family vs. individual level leads to a downward estimate of heritability compared to the model that uses individuals.
4) **Model 1 vs Model 3 vs Model 6** - Inclusion of all effects including sire (model 1), all effects minus sire (model 6), and the simpler model including animal, dam, and jar (model 3) appear to generate similar estimates of heritability.

# Variance and Heritability Comparisons for "Full Model" (Model 6) and "Best Model" (Model 3)
```{r echo=F}
sumTabSub <- subset(sumTab,sumTab$model == "Model3" | sumTab$model == "Model6")

model <- c(sumTabSub$model,sumTabSub$model)
treatment <- paste0("P-",sumTabSub$Parent_Env," O-",sumTabSub$Offspring_Env)
treatment <- c(treatment,treatment)
varComponent <- c(rep("genetic",times=nrow(sumTabSub)),rep("dam",times=nrow(sumTabSub)))
meanH <- c(sumTabSub$animal_H2_mean,sumTabSub$dam_H2_mean)
lowerH <- c(sumTabSub$animal_H2_lower,sumTabSub$dam_H2_lower)
upperH <- c(sumTabSub$animal_H2_upper,sumTabSub$dam_H2_upper)

df <- data.frame(model,treatment,varComponent,meanH,lowerH,upperH)

ggplot(df,aes(x=treatment,y=meanH)) +
  facet_grid(cols=vars(varComponent),rows=vars(model)) +
  ylim(0,1) +
  geom_point() +
  geom_errorbar(aes(ymin=lowerH, ymax=upperH)) +
  labs(y="Proportion of Variation",x="") +
  theme_cowplot() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 25, vjust = 0.5, hjust=1))
```

### Take-aways

1) Large credibility intervals and low heritability for all parent and offspring treatment combinations.
2) **P-Exposed O-Exposed vs others** - Heritability seems slightly higher for the ExposedxExposed scenario, but not significantly so based on the credibility intervals.
3) **dam vs genetic** - Generally, the maternal component is larger in in larvae from parents in ambient conditions, whereas the genetic component appears larger in larvae from exposed parents.

### Variance Components for Model 3
```{r echo=F,fig.height=10,warning=F}
model_var <- function(x,title){
  ## Plot for visualizing all random variance components
ref <- data.frame(x)
rdf <- melt(ref,value.name="Area") # Replace growth rate is using another phenotype
density_plot <- ggplot(rdf,aes(x=sqrt(Area),color=variable)) +
  xlim(0,1) +
  geom_density() +
  labs(x="",title=title) +
  theme_cowplot() +
  theme(legend.position = "none",plot.title = element_text(hjust = 0.5))
boxplot_plot <- ggplot(rdf,aes(x=sqrt(Area),y=variable,color=variable)) +
  xlim(0,1) +
  geom_boxplot() +
  labs(x="Larval Area Components of Variation (SD)") +
  theme_cowplot() +
  theme(legend.position = "none")
p1 <- plot_grid(density_plot,boxplot_plot,ncol = 1,align="v")
p1
}

model_var(mod_output$cc$mod3$model$VCV,"Parent:Control Offspring:Control")
model_var(mod_output$ce$mod3$model$VCV,"Parent:Control Offspring:Exposed")
model_var(mod_output$ec$mod3$model$VCV,"Parent:Exposed Offspring:Control")
model_var(mod_output$ee$mod3$model$VCV,"Parent:Exposed Offspring:Exposed")
```

### Take-aways 

1) **Large residual variation** : We can see that residual variation is large compared to all other variance components.
2) **Wide distributions for animal and dam** : In all cases the model struggles to consistently estimate the variance explained by animal and dam, leading to the broad distributions relative to the other variance components.

```{r echo=F,fig.width=10,fig.height=10,warning=F}
model_2d <- function(x,title){
  ref <- data.frame(x)
  ggplot(ref, aes(x=sqrt(animal),y=sqrt(damID))) + 
  xlim(0,1) + ylim(0,1) +
  geom_point(alpha=0.5) +
  labs(x="Animal Variance Component (SD)",y="Dam Variance Component (SD)",title=title) + 
  geom_density2d() + 
  geom_abline(intercept = mean(sqrt(ref$damID)),slope=0,color="blue",size=1.5) +
  geom_vline(xintercept=mean(sqrt(ref$animal)),color="green",size=1.5) +
  geom_smooth(method='lm',colour="orange",size=1.5) +
  geom_abline(int=0,slope=1,size=1.5) +
  theme_cowplot() + 
    theme(plot.title = element_text(hjust = 0.5))
}

p1 <- model_2d(mod_output$cc$mod3$model$VCV,"Parent:Control Offspring:Control")
p2 <- model_2d(mod_output$ce$mod3$model$VCV,"Parent:Control Offspring:Exposed")
p3 <- model_2d(mod_output$ec$mod3$model$VCV,"Parent:Exposed Offspring:Control")
p4 <- model_2d(mod_output$ee$mod3$model$VCV,"Parent:Exposed Offspring:Exposed")

plot_grid(p1,p2,p3,p4,nrow=2)
```
**Figure Caption**: Black line is 1:1 line, blue line is the mean maternal SD (y-axis), green is mean genetic SD (x-axis), orange is linear model fit using ```geom_smooth``` 

### Take-aways

1) Does appear to be a slightly negative relationship between the maternal and genetic variance components. This is indicates that perhaps part of the reason the distributions of the SDs for this two components are wide is partially due to the model struggling to handle the correlation between these two components?? Although the outliers SD estimate for dam (large values on the y-axis) do not appear to necessarily lead to a dramatic decrease in the genetic estimate.

# Main Observations

1) Hard to draw many conclusions from the model. Mean heritability is low for all comparisons, but given the large credibility intervals around most of the variance components it is hard to draw conclusions.
2) There isn't a clearly dominant variance component (e.g. genetic, maternal, or experimental), and it looks like the model struggles to consistently converge on estimates for each of the components given the large credibility intervals.
3) Looking at the model diagnostics it looks like including Seatable is problematic, given the posterior distribution, although not including it in the model doesn't appear to impact heritability estimates much. As a result it seems like using a model with animal, dam, and jar may be the simplest, best model in this case (Model 3).

# Future Directions

1) Finish reading through Hadley literature to identify an other things to consider when parameterizing the model. Specifically, is there a way to improve the posterior distributions for animal and dam effects (so they appear more normally distributed).
2) Add measure of parental methylation (when we have it).
3) **Extend Goal** : Estimate parental relatedness and use kinship matrix rather than traditional pedigree to improve our ability to estimate heritability (right now the assumption is that all parents are unrelated, the first step is to check how true this is).


# Model Diagnostics

## Model 3 (Partial Model)
```{r echo=F}
## Diagnostics
m1 <- mod_output$cc$mod3
m2 <- mod_output$ce$mod3
m3 <- mod_output$ec$mod3
m4 <- mod_output$ee$mod3
```

### Data : Parent Control - Offspring Control
```{r echo=F}
m1$model_summary
plot(m1$model$Sol,ask=F)
autocorr.diag(m1$model$Sol)
heidel.diag(m1$model$Sol)
```

```{r echo=F,fig.height=12,fig.width=10}
plot(m1$model$VCV,ask=F)
autocorr.diag(m1$model$VCV)
heidel.diag(m1$model$VCV)
```

### Data : Parent Control - Offspring Exposed
```{r echo=F}
m2$model_summary
plot(m2$model$Sol,ask=F)
autocorr.diag(m2$model$Sol)
heidel.diag(m2$model$Sol)
```

```{r echo=F,fig.height=12,fig.width=10}
plot(m2$model$VCV,ask=F)
autocorr.diag(m2$model$VCV)
heidel.diag(m2$model$VCV)
```

### Data : Parent Exposed - Offspring Control
```{r echo=F}
m3$model_summary
plot(m3$model$Sol,ask=F)
autocorr.diag(m3$model$Sol)
heidel.diag(m3$model$Sol)
```

```{r echo=F,fig.height=12,fig.width=10}
plot(m3$model$VCV,ask=F)
autocorr.diag(m3$model$VCV)
heidel.diag(m3$model$VCV)
```

### Data : Parent Exposed - Offspring Exposed
```{r echo=F}
m4$model_summary
plot(m4$model$Sol,ask=F)
autocorr.diag(m4$model$Sol)
heidel.diag(m4$model$Sol)
```

```{r echo=F,fig.height=12,fig.width=10}
plot(m4$model$VCV,ask=F)
autocorr.diag(m4$model$VCV)
heidel.diag(m4$model$VCV)
```

## Model 6 (Full Model)
```{r echo=F}
## Diagnostics
m1 <- mod_output$cc$mod7
m2 <- mod_output$ce$mod7
m3 <- mod_output$ec$mod7
m4 <- mod_output$ee$mod7
```

### Data : Parent Control - Offspring Control
```{r echo=F}
m1$model_summary
plot(m1$model$Sol,ask=F)
autocorr.diag(m1$model$Sol)
heidel.diag(m1$model$Sol)
```

```{r echo=F,fig.height=12,fig.width=10}
plot(m1$model$VCV,ask=F)
autocorr.diag(m1$model$VCV)
heidel.diag(m1$model$VCV)
```

### Data : Parent Control - Offspring Exposed
```{r echo=F}
m2$model_summary
plot(m2$model$Sol,ask=F)
autocorr.diag(m2$model$Sol)
heidel.diag(m2$model$Sol)
```

```{r echo=F,fig.height=12,fig.width=10}
plot(m2$model$VCV,ask=F)
autocorr.diag(m2$model$VCV)
heidel.diag(m2$model$VCV)
```

### Data : Parent Exposed - Offspring Control
```{r echo=F}
m3$model_summary
plot(m3$model$Sol,ask=F)
autocorr.diag(m3$model$Sol)
heidel.diag(m3$model$Sol)
```

```{r echo=F,fig.height=12,fig.width=10}
plot(m3$model$VCV,ask=F)
autocorr.diag(m3$model$VCV)
heidel.diag(m3$model$VCV)
```

### Data : Parent Exposed - Offspring Exposed
```{r echo=F}
m4$model_summary
plot(m4$model$Sol,ask=F)
autocorr.diag(m4$model$Sol)
heidel.diag(m4$model$Sol)
```

```{r echo=F,fig.height=12,fig.width=10}
plot(m4$model$VCV,ask=F)
autocorr.diag(m4$model$VCV)
heidel.diag(m4$model$VCV)
```

