  ---
title: "Heritability Growth Estimate"
output: 
  github_document:
    toc : true
  toc_depth: 3
  theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include=FALSE}
library(MCMCglmm,verbose = F)
library(ggplot2,verbose = F)
library(ggpubr,verbose = F)
library(cowplot,verbose = F)
library(dplyr,verbose = F)
library(knitr,verbose = F)
library(brms,verbose = F)
library(rstan,verbose = F)
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
```

## Overview 

In this markdown I use an animal model to estimate narrow-sense heritability and a measure of maternal inheritance. In both cases these estimates were measured as:

![](http://www.sciweavers.org/tex2img.php?eq=%5Cfrac%7B%5Csigma_%7Ba%7D%7D%7B%5Csigma_%7BP%7D%7D&bc=White&fc=Black&im=png&fs=12&ff=arev&edit=0)

Where &sigma;<sub>a</sub> is the additive genetic variance (or variance associated with `damID`) and &sigma;<sub>P</sub> is the total phenotypic variance (calculated by summing all sources of estimated variance).

I explore 5 different models that examine the inclusion of different random effects, priors, and input data (individual vs. mean family summary by jar). Table of models:

| Model | Data Input |Prior Type | Prior Name | Random Effects |
|:-----:|:----------:|:---------:|:----------:|:--------------:|
| Model 1 | Individual | Chi | Prior2.5 | ~ animal + DamID + SireID + JarTrt + SeaTable |
| Model 2 | Individual | Chi | Prior2.3 | ~ animal + DamID + JarTrt |
| Model 3 | Individual | Inverse-Gamma | Prior1.3 | ~ animal + DamID + JarTrt |
| Model 4 | Mean Family | Chi | Prior2.3 | ~ animal + DamID + SeaTable |
| Model 5 | Mean Family | Chi | Prior2.2 | ~ animal + DamID |

* **Model 1** : Complete model with all random effects. 
* **Model 2** : Model with Sire and SeaTable removed. This was done because Sire explained very littl variation and SeaTable was also largely uninformative, but would occasionaly generate a very large estimate. I am not sure what is generating this pattern, but it caused the mean phenotypic variance to become greatly inflated.
* **Model 3** : Model with reduced random effects and uninformative prior (inverse-gamma)
* **Model 4** : Model using growth data summarized (mean) for each Jar. This greatly reduced replication, but I was interested in seeing how it impacted heritability estimates compared to the more traditional animal model. In this model i included all possible random effects (no Jar).
* **Model 5** : Continuation of Model 4 but without SeaTable as random effect (same rationale as Model 2).

## Data
```{r eval=F}
jr <- readRDS("~/Github/2018OAExp_larvae/input_files/JarHeritabilityData.RDS")
ind <- readRDS("~/Github/2018OAExp_larvae/input_files/IndHeritabilityData.RDS")

pheno <- ind$pheno
hist(pheno$GrowthPerDay)
pheno$GrowthScale <-scale(pheno$GrowthPerDay)
ped1 <- jr$ped
jrp <- jr$pheno
jrp$GrowthScale <-scale(jrp$Growth)

# Control Parents - Control Offspring 
cc <- subset(pheno, pheno$JarTrt == "Control" & pheno$ParentTrt == 400)
dam_cc <- data.frame(id=unique(cc$damID),dam=NA,sire=NA)
sire_cc <- data.frame(id=unique(cc$sireID),dam=NA,sire=NA)
ped_cc <- rbind(dam_cc,sire_cc,data.frame(id=cc$animal,dam=cc$damID,sire=cc$sireID))

ped_cc_brms <- ped_cc
ped_cc_brms$dam[ped_cc$dam == 0] <- NA
ped_cc_brms$sire[ped_cc$sire == 0] <- NA

jr_cc <- subset(jrp, jrp$JarTrt == "Control" & jrp$ParentTrt == 400)
dam_jr_cc <- data.frame(id=unique(jr_cc$damID),dam=NA,sire=NA)
sire_jr_cc <- data.frame(id=unique(jr_cc$sireID),dam=NA,sire=NA)
ped_jr_cc <- rbind(dam_jr_cc,sire_jr_cc,data.frame(id=jr_cc$animal,dam=jr_cc$damID,sire=jr_cc$sireID))

# Control Parents - Exposed Offspring 
ce <- subset(pheno, pheno$JarTrt == "Exposed" & pheno$ParentTrt == 400)
dam_ce <- data.frame(id=unique(ce$damID),dam=NA,sire=NA)
sire_ce <- data.frame(id=unique(ce$sireID),dam=NA,sire=NA)
ped_ce <- rbind(dam_ce,sire_ce,data.frame(id=ce$animal,dam=ce$damID,sire=ce$sireID))

ped_ce_brms <- ped_ce
ped_ce_brms$dam[ped_ce$dam == 0] <- NA
ped_ce_brms$sire[ped_ce$sire == 0] <- NA

jr_ce <- subset(jrp, jrp$JarTrt == "Exposed" & jrp$ParentTrt == 400)
dam_jr_ce <- data.frame(id=unique(jr_ce$damID),dam=NA,sire=NA)
sire_jr_ce <- data.frame(id=unique(jr_ce$sireID),dam=NA,sire=NA)
ped_jr_ce <- rbind(dam_jr_ce,sire_jr_ce,data.frame(id=jr_ce$animal,dam=jr_ce$damID,sire=jr_ce$sireID))

# Exposed Parents - Control Offspring 
ec <- subset(pheno, pheno$JarTrt == "Control" & pheno$ParentTrt == 2600)
dam_ec <- data.frame(id=unique(ec$damID),dam=NA,sire=NA)
sire_ec <- data.frame(id=unique(ec$sireID),dam=NA,sire=NA)
ped_ec <- rbind(dam_ec,sire_ec,data.frame(id=ec$animal,dam=ec$damID,sire=ec$sireID))

ped_ec_brms <- ped_ec
ped_ec_brms$dam[ped_ec$dam == 0] <- NA
ped_ec_brms$sire[ped_ec$sire == 0] <- NA

jr_ec <- subset(jrp, jrp$JarTrt == "Control" & jrp$ParentTrt == 2600)
dam_jr_ec <- data.frame(id=unique(jr_ec$damID),dam=NA,sire=NA)
sire_jr_ec <- data.frame(id=unique(jr_ec$sireID),dam=NA,sire=NA)
ped_jr_ec <- rbind(dam_jr_ec,sire_jr_ec,data.frame(id=jr_ec$animal,dam=jr_ec$damID,sire=jr_ec$sireID))

# Exposed Parents - Exposed Offspring 
ee <- subset(pheno, pheno$JarTrt == "Exposed" & pheno$ParentTrt == 2600)
dam_ee <- data.frame(id=unique(ee$damID),dam=NA,sire=NA)
sire_ee <- data.frame(id=unique(ee$sireID),dam=NA,sire=NA)
ped_ee <- rbind(dam_ee,sire_ee,data.frame(id=ee$animal,dam=ee$damID,sire=ee$sireID))

ped_ee_brms <- ped_ee
ped_ee_brms$dam[ped_ee$dam == 0] <- NA
ped_ee_brms$sire[ped_ee$sire == 0] <- NA

jr_ee <- subset(jrp, jrp$JarTrt == "Exposed" & jrp$ParentTrt == 2600)
dam_jr_ee <- data.frame(id=unique(jr_ee$damID),dam=NA,sire=NA)
sire_jr_ee <- data.frame(id=unique(jr_ee$sireID),dam=NA,sire=NA)
ped_jr_ee <- rbind(dam_jr_ee,sire_jr_ee,data.frame(id=jr_ee$animal,dam=jr_ee$damID,sire=jr_ee$sireID))
```

## Priors
```{r eval=F}
# Inverse-Gamma (uninformative)
prior1.3 <- list(R = list(V=1, nu=0.002), G = list(G1 = list(V=1, nu=0.002),
                                                   G2 = list(V=1, nu=0.002),
                                                   G3 = list(V=1, nu=0.002)))
# Chi-Square (mildly informative)
prior2.2<-list(R=list(V=1,nu=1),
               G=list(G1=list(V=1,nu=1,alpha.mu=0,alpha.V=1000),
                      G2=list(V=1,nu=1,alpha.mu=0,alpha.V=1000)))

prior2.3<-list(R=list(V=1,nu=1),
               G=list(G1=list(V=1,nu=1,alpha.mu=0,alpha.V=1000),
                      G2=list(V=1,nu=1,alpha.mu=0,alpha.V=1000),
                      G3=list(V=1,nu=1,alpha.mu=0,alpha.V=1000)))

prior2.5<-list(R=list(V=1,nu=1),
               G=list(G1=list(V=1,nu=1,alpha.mu=0,alpha.V=1000),
                      G2=list(V=1,nu=1,alpha.mu=0,alpha.V=1000),
                      G3=list(V=1,nu=1,alpha.mu=0,alpha.V=1000),
                      G4=list(V=1,nu=1,alpha.mu=0,alpha.V=1000),
                      G5=list(V=1,nu=1,alpha.mu=0,alpha.V=1000)))
```

## Model Function
```{r eval=F}
runModel <- function(pheno,ped,prior,model_name,parentTrt,offspringTrt){
  if(model_name == "Model1"){
    model_mcmc  <- MCMCglmm(GrowthScale ~ 1,
                            random = ~ animal + damID + sireID + JarID + JarSeatable,
                            family = "gaussian",
                            prior = prior,
                            pedigree = ped,
                            data = pheno,
                            nitt = 5000000,
                            burnin = 10000,
                            thin = 1000)
  }
  if(model_name == "Model2"){
    model_mcmc  <- MCMCglmm(GrowthScale ~ 1,
                             random = ~ animal + damID + JarID ,
                             family = "gaussian",
                             prior = prior,
                             pedigree = ped,
                             data = pheno,
                             nitt = 1000000,
                             burnin = 10000,
                             thin = 1000)
  }
  if(model_name == "Model3"){
    model_mcmc  <- MCMCglmm(GrowthScale ~ 1,
                             random = ~ animal + damID + JarID,
                             family = "gaussian",
                             prior = prior,
                             pedigree = ped,
                             data = pheno,
                             nitt = 5000000,
                             burnin = 10000,
                             thin = 1000)
  }
  if(model_name == "Model4"){
    model_mcmc  <- MCMCglmm(GrowthScale ~ 1,
                             random = ~ animal + damID + SeaTable,
                             family = "gaussian",
                             prior = prior,
                             pedigree = ped,
                             data = pheno,
                             nitt = 5000000,
                             burnin = 10000,
                             thin = 1000)
  }
  if(model_name == "Model5"){
    model_mcmc  <- MCMCglmm(GrowthScale ~ 1,
                             random = ~ animal + damID,
                             family = "gaussian",
                             prior = prior,
                             pedigree = ped,
                             data = pheno,
                             nitt = 5000000,
                             burnin = 10000,
                             thin = 1000)
  }
  
  model_sum <- summary(model_mcmc)
  list_var <- list()
  for(i in 1:ncol(model_mcmc$VCV)){
    list_var[[i]] <- model_mcmc$VCV[,colnames(model_mcmc$VCV)[i]]
  }
  names(list_var)<-colnames(model_mcmc$VCV)
  mt1 <- data.frame(matrix(unlist(list_var), ncol = length(colnames(model_mcmc$VCV)), byrow = F))
  colnames(mt1)<-colnames(model_mcmc$VCV)
  mt1$Pheno <- rowSums(mt1)
  mt1$animal_H2 <- mt1$animal / mt1$Pheno
  mt1$damID <- mt1$damID / mt1$Pheno
  
  eff <- c("animal","damID","sireID","JarID","JarSeatable","Pheno")
  values <- NULL
  for(i in eff){
    if(!is.na(match(i,colnames(mt1)))){
      values <- c(values,mean(mt1[,i]))
      values <- c(values,quantile(mt1[,i],0.025))
      values <- c(values,quantile(mt1[,i],0.975))
    }else{
      values <- c(values,NA)
      values <- c(values,NA)
      values <- c(values,NA)
    }
  }
  col_names <- paste0(rep(eff,each=3),c("_mean","_lower","_upper"))
  values <- data.frame(round(t(as.matrix(values)),5))
  colnames(values) <- col_names
  summary_table <- data.frame(model=model_name,Parent_Env=parentTrt,Offspring_Env=offspringTrt,values)
  
  return(list(model=model_mcmc,model_summary=model_sum,variance_output=mt1,summary_table))
}
```

## Model Runs
```{r eval=F}
# Control - Control
cc_mod1 <- runModel(cc,ped_cc,prior2.2,"Model1","Control","Control")
cc_mod2 <- runModel(cc,ped_cc,prior2.2,"Model2","Control","Control")
cc_mod3 <- runModel(cc,ped_cc,prior2.2,"Model3","Control","Control")
cc_mod4 <- runModel(jr_cc,ped_jr_cc,prior2.2,"Model4","Control","Control")
cc_mod5 <- runModel(jr_cc,ped_jr_cc,prior2.2,"Model5","Control","Control")
cc <- list(mod1=cc_mod1,mod2=cc_mod1,mod3=cc_mod1,mod4=cc_mod1,mod5=cc_mod1) 
# Control - Exposed
ce_mod1 <- runModel(ce,ped_ce,prior2.2,"Model1","Control","Exposed")
ce_mod2 <- runModel(ce,ped_ce,prior2.2,"Model2","Control","Exposed")
ce_mod3 <- runModel(ce,ped_ce,prior2.2,"Model3","Control","Exposed")
ce_mod4 <- runModel(jr_ce,ped_jr_ce,prior2.2,"Model4","Control","Exposed")
ce_mod5 <- runModel(jr_ce,ped_jr_ce,prior2.2,"Model5","Control","Exposed")
ce <- list(mod1=ce_mod1,mod2=ce_mod1,mod3=ce_mod1,mod4=ce_mod1,mod5=ce_mod1) 
# Exposed - Control
ec_mod1 <- runModel(ec,ped_ec,prior2.2,"Model1","Exposed","Control")
ec_mod2 <- runModel(ec,ped_ec,prior2.2,"Model2","Exposed","Control")
ec_mod3 <- runModel(ec,ped_ec,prior2.2,"Model3","Exposed","Control")
ec_mod4 <- runModel(jr_ec,ped_jr_ec,prior2.2,"Model4","Exposed","Control")
ec_mod5 <- runModel(jr_ec,ped_jr_ec,prior2.2,"Model5","Exposed","Control")
ec <- list(mod1=ec_mod1,mod2=ec_mod1,mod3=ec_mod1,mod4=ec_mod1,mod5=ec_mod1) 
# Exposed - Exposed
ee_mod1 <- runModel(ee,ped_ee,prior2.2,"Model1","Exposed","Exposed")
ee_mod2 <- runModel(ee,ped_ee,prior2.2,"Model2","Exposed","Exposed")
ee_mod3 <- runModel(ee,ped_ee,prior2.2,"Model3","Exposed","Exposed")
ee_mod4 <- runModel(jr_ee,ped_jr_ee,prior2.2,"Model4","Exposed","Exposed")
ee_mod5 <- runModel(jr_ee,ped_jr_ee,prior2.2,"Model5","Exposed","Exposed")
ee <- list(mod1=ee_mod1,mod2=ee_mod1,mod3=ee_mod1,mod4=ee_mod1,mod5=ee_mod1)

mod_output <- list(cc=cc,ce=ce,ec=ec,ee=ee)
```

```{r echo=F}
mod_output <- readRDS("~/Desktop/GrowthHeritabilityEstimate.RDS")

subchunkify <- function(g, fig_height=7, fig_width=5) {
  g_deparsed <- paste0(deparse(
    function() {g}
  ), collapse = '')
  
  sub_chunk <- paste0("
  `","``{r sub_chunk_", floor(runif(1) * 10000), ", fig.height=", fig_height, ", fig.width=", fig_width, ", echo=FALSE}",
  "\n(", 
    g_deparsed
    , ")()",
  "\n`","``
  ")
  
  cat(knitr::knit(text = knitr::knit_expand(text = sub_chunk), quiet = TRUE))
}
```

## Diagnostic Outputs
```{r fig.width=10,fig.height=8,echo=FALSE}
sumTab <- NULL 
## Diagnostics
for(i in mod_output){
  for(j in i){
    print(" ")
    print(paste("#########################    ",j$summary[,1:3],"    #########################"))
    print(" ")
    sumTab <- rbind(sumTab,j$summary)
    print(j$model_summary)
    print(" ")
    # Plot of posteriors
    plot(j$model$Sol,ask=F)
    print(" ")
    plot(j$model$VCV,ask=F)
    print(" ")
    # Autocorrelation assessment
    print(autocorr.diag(j$model$Sol))
    print(" ")
    print(autocorr.diag(j$model$VCV))
    print(" ")
    # Effective sample size
    print(effectiveSize(j$model$Sol))
    print(" ")
    print(effectiveSize(j$model$VCV))
    print(" ")
    # Test of model convergence
    print(heidel.diag(j$model$VCV))
    print(" ")
    print(" ")
    print(" ")
  }
}
```

## Summary Tables

### Model Information (same table as above)

| Model | Data Input |Prior Type | Prior Name | Random Effects |
|:-----:|:----------:|:---------:|:----------:|:--------------:|
| Model 1 | Individual | Chi | Prior2.5 | ~ animal + DamID + SireID + JarTrt + SeaTable |
| Model 2 | Individual | Chi | Prior2.3 | ~ animal + DamID + JarTrt |
| Model 3 | Individual | Inverse-Gamma | Prior1.3 | ~ animal + DamID + JarTrt |
| Model 4 | Mean Family | Chi | Prior2.3 | ~ animal + DamID + SeaTable |
| Model 5 | Mean Family | Chi | Prior2.2 | ~ animal + DamID |

### Mode Outputs
```{r echo=F}
kable(sumTab)
```

## Summary Plot
```{r echo=F, fig.width=10,fig.height=10}
animal_plot <- ggplot(sumTab,aes(x=interaction(Parent_Env,Offspring_Env,model),
                  y=animal_H2_mean,
                  group=interaction(model,Parent_Env,Offspring_Env),
                  #linetype=model,
                  colour=interaction(Parent_Env,Offspring_Env))) + 
  geom_point() + 
  geom_errorbar(aes(ymin=animal_H2_lower,ymax=animal_H2_upper)) +
  theme_cowplot() + 
  #annotate(geom = "text", x = seq_len(nrow(sumTab)), y = -0.05, label = rep(unique(sumTab$model),each=4), size = 4,angle=45) +
  annotate(geom = "text", x = 2.5+4*c(0:4), y = -0.1, label = unique(sumTab$model), size = 6) +
    coord_cartesian(ylim = c(-0.02,1), expand = FALSE, clip = "off") +
  labs(y="Heritability",title="Animal",colour="Parent.Offspring \n    Environment") +
  theme(plot.margin = unit(c(1, 1, 4, 1), "lines"),
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank())

dam_plot <- ggplot(sumTab,aes(x=interaction(Parent_Env,Offspring_Env,model),
                                 y=dam_H2_mean,
                                 group=interaction(model,Parent_Env,Offspring_Env),
                                 #linetype=model,
                                 colour=interaction(Parent_Env,Offspring_Env))) + 
  geom_point() + 
  geom_errorbar(aes(ymin=dam_H2_lower,ymax=dam_H2_upper)) +
  theme_cowplot() + 
  #annotate(geom = "text", x = seq_len(nrow(sumTab)), y = -0.05, label = rep(unique(sumTab$model),each=4), size = 4,angle=45) +
  annotate(geom = "text", x = 2.5+4*c(0:4), y = -0.1, label = unique(sumTab$model), size = 6) +
  coord_cartesian(ylim = c(-0.02,1), expand = FALSE, clip = "off") +
  labs(y="Heritability",title="Dam",colour="Parent.Offspring \n    Environment") +
  theme(plot.margin = unit(c(1, 1, 4, 1), "lines"),
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank())

plot_grid(animal_plot,dam_plot,nrow=2)
```
