---
title: "Estimating plasticity of larvae responses to OA"
output: 
  github_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=F}
library(MCMCglmm)
library(ggplot2)
library(ggpubr)
library(cowplot)
library(dplyr)
library(reshape2)
```

## Overview  
With the 2018 larval data I think we have the potential to estimate both traditional heritability of phenotypic variance and heritability of plasticity. I haven't seen this frequently discussed in the literature from the perspective of heritability, but thinking of plasticity as the slope of the reaction norm is common in other areas of evolutionary biology. Given that we split individuals from each parental cross into both larval control and exposed environments, we should have a simple two environment reaction norm for each cross. Below I discuss two approaches for slope estimation, which could subsequently be used as a response variable with an animal model to estimate heritability of plasticity. 


**Larval Data**
```{r}
jr <- readRDS("~/Github/2018OAExp_larvae/input_files/JarHeritabilityData.RDS")
ind <- readRDS("~/Github/2018OAExp_larvae/input_files/IndHeritabilityData.RDS")
```

```{r include=F}
jrp <- jr$pheno
jrp$parentalComb <- paste(jrp$damID,jrp$sireID,sep="_")
jrp %>% group_by(parentalComb,JarTrt) %>% summarise(ParentTrt=unique(ParentTrt),Growth=mean(Growth)) -> jrp_summary
jrp %>% group_by(parentalComb) %>% summarise(ParentTrt=unique(ParentTrt)) -> jrp_pcombo
```

### Initial Visualization of Data

Here we the estimated growth for each jar by larval treatment (x-axis) and facet by parental treatment. You can see that OA has a significant effect on larvae growth and parental treatment also seems to impact growth in both treatments.

```{r}
ggplot(jrp,aes(x=JarTrt,y=Growth,colour=JarTrt)) + 
  facet_grid(.~ParentTrt) + 
  geom_boxplot() +
  theme_cowplot() + 
  labs(x="Larval Environment",y="Larval Growth","Growth by Jar")
```

Next, we summarize the mean growth among replicate jars (n approx. = 3) and use grey lines to show the phenotypic plasticity (slope) of each family cross. This was possible because zygotes created from parental crosses were split with half going into a larval control treatment and half going into an exposed treatment.

```{r}
ggpaired(jrp_summary, x = "JarTrt", y = "Growth",id="parentalComb",
         color = "JarTrt", line.color = "gray", line.size = 0.4,facet.by = "ParentTrt") + 
  labs(x = "Larval Environment" , y = "Larval Growth") +
  theme(legend.position = "none")
```

I think it makes sense to think of these grey lines as family specific reaction norms, with the slope of each family representing a measure of phenotypic plasticity of a particular phenotype (e.g. growth) in response to a specific environmental stress (e.g. OA). In turn, to examine the heritability of plasticity it might be appropriate to estimate this slope for each family, and use this as our response variable in the animal model.

Unfortunately, I have not see much literature on this approach, but I can think of two ways in which we can estimate the slope for each cross:  
* **(Option 1)** : Use linear regresion to estimate the slope for each family.  
* **(Option 2)** : Use bootstrapping to estimate a slope distribution for each family.  

The former option is the simplest, but may be problematic because we have a small number of replicates (n=3) and our data may not meet the assumptions of regression. Alternatively, bootstrapping has fewer assumptions, but generates a distribution rather than a single slope estimate. I think we might then have to generate a number of test datasets in which we randomly draw a single slope value for each family from their respective distributions, then run each of these datasets through the animal model to generate a distribution of possible mean heritability estimates.

## Option 1 -  Estimating slopes via regression

Here we fit a linear reaction norm to each family using regression. To this I simply fitted a simple linear model (`lm`) to each family separately (n = ~6 (3 control and 3 exposed)) using phenotype as the response variable and jar pH as the explainatory variable. The slope from this model represents an approximate measure of plasticity for each family.  

```{r warning=F}
y <- c("Growth","Area","Diameter","Perimeter","MMR","PDR","LarvaeSurvived","AbnormalCount","PercentCilia")
x <- "JarpHNBS"
slope_mat <- matrix(ncol=length(y),nrow=length(unique(jrp$parentalComb)))
colnames(slope_mat) <- y
intercept_mat <- matrix(ncol=length(y),nrow=length(unique(jrp$parentalComb)))
colnames(intercept_mat) <- y
for( i in 1:length(unique(jrp$parentalComb))){
  for(j in 1:length(y)){
    temp <- jrp[jrp$parentalComb == unique(jrp$parentalComb)[i],]
    lm_out <- lm(temp[,y[j]]~temp[,x])
    lm_sum <- summary(lm_out)
    intercept_mat[i,j] <- lm_sum$coefficients[1,1]
    slope_mat[i,j] <- lm_sum$coefficients[2,1]
  }
}

jar_growth <- data.frame(parentalComb=unique(jrp$parentalComb))
jar_growth <- left_join(jar_growth,jrp_pcombo,by="parentalComb")
jar_intercept <- data.frame(jar_growth,intercept_mat)
jar_slope <- data.frame(jar_growth,slope_mat)

jar <- list(intercept=jar_intercept,slope=jar_slope)
```

```{r echo=F}
plot_slopes <- list()
plot_intercepts <- list()
for( i in 1:length(y)){
  temp <- data.frame(Parent_Treatment=as.factor(jar$slope$ParentTrt),Slope=jar$slope[,i+2])
  out <- ggplot(temp,aes(Slope,group=Parent_Treatment,colour=Parent_Treatment)) + 
    geom_density() +
    theme_cowplot() +
    labs(title=y[i],x="",y="") +
    theme(legend.position = "none",
          plot.title = element_text(hjust = 0.5))
  plot_slopes[[i]] <- out
  temp <- data.frame(Parent_Treatment=as.factor(jar$intercept$ParentTrt),Intercept=jar$intercept[,i+2])
  out <- ggplot(temp,aes(Intercept,group=Parent_Treatment,colour=Parent_Treatment)) + 
    geom_density() +
    theme_cowplot() +
    labs(title=y[i],x="",y="") +
    theme(legend.position = "none",
          plot.title = element_text(hjust = 0.5))
  plot_intercepts[[i]] <- out
    
}
```

### Distribution of slopes for various phenotypes by parental treatment  
```{r echo=F,fig.width=10,fig.height=10}
plot_grid(plotlist=plot_slopes)
```

## Option 2 - Use bootstraping to generate a distribution of slopes based on the average of randomly sampled test datasets  

In this option, rather than estimating the slope for a family using regression I make a list of all possible slopes given every combination of control and exposed replicates per family. For example, for family CF02_CM02 I would calculate the slope of Control_Rep1 x Exposed_Rep2, Control_Rep1 x Exposed_Rep2, and Control_Rep1 x Exposed_Rep3, then do the same for Control_Rep2 and Control_Rep3. This would generate a list with 9 possible slopes, based on all possible replicate combinations. After doing this for all familys I then sample one slope from each family to create a sample dataset. I do this 1 thousand times (`REP=1000`). I am left with 1000 thousand test datasets for each parental treatment, for each phenotype. I visuallize this data below by plotting the distribution of slopes summarized across all replicate datasets.

```{r}
y <- c("Growth","Area","Diameter","Perimeter","MMR","PDR","LarvaeSurvived","PercentCilia")
x <- "JarpHNBS"
slope_comb <- rep(list(NULL),times=length(y))
names(slope_comb) <- y
jrp <- jrp[jrp$JarID != "J279_B1",]
jrp$AbnormalCount <- as.numeric(jrp$AbnormalCount)
jrp %>% group_by(parentalComb) %>% summarise(ParentTrt=unique(ParentTrt)) -> jrp_pcombo

for(i in 1:length(y)){
  for(j in 1:length(unique(jrp$parentalComb))){
    temp <- jrp[jrp$parentalComb == unique(jrp$parentalComb)[j],]
    tc_y <- temp[temp$JarTrt == "Control",y[i]]
    tc_x <- temp[temp$JarTrt == "Control",x]
    te_y <- temp[temp$JarTrt == "Exposed",y[i]]
    te_x <- temp[temp$JarTrt == "Exposed",x]
    temp_slope <- NULL
    for(m in 1:length(tc_y)){
      for(n in 1:length(te_y)){
        temp_slope <- c(temp_slope,c(tc_y[m]-te_y[n])/c(tc_x[m]-te_x[n]))
      }
    }
    if(j == 1){
      slope_comb[[y[i]]] <- list(name=temp_slope)
    }else{
      slope_comb[[y[i]]] <- c(slope_comb[[y[i]]],name=list(temp_slope))
    }
  }
  names(slope_comb[[i]]) <- unique(jrp$parentalComb)
}
```

```{r}
REP=1000
slope_reps <- rep(list(NULL),times=length(slope_comb))
names(slope_reps) <- names(slope_comb)
for(j in 1:length(slope_reps)){
  temp <- slope_comb[[names(slope_comb)[j]]]
  temp_mat <- matrix(nrow=(length(temp)),ncol=REP)
  colnames(temp_mat) <- paste0("Rep_",c(1:REP))
  rownames(temp_mat) <- names(temp)
  for(i in 1:length(temp)){
    temp_list <- temp[[i]]
    temp_mat[i,]<- sample(temp_list,size=REP,replace = T)
  }
  slope_reps[[j]] <- temp_mat
}
```

```{r echo=F,fig.width=10,fig.height=10}
plot_slope_list2 <- list()
for(i in 1:length(slope_reps)){
  out <- data.frame(family=rownames(slope_reps[[names(slope_reps[i])]]),Parent_Trt=as.factor(jrp_pcombo$ParentTrt),slope_reps[[names(slope_reps)[i]]])
  out2 <- melt(out,id.vars = c("family","Parent_Trt"))
  plot_slope_list2[[i]] <- ggplot(out2,aes(value,group=Parent_Trt,colour=Parent_Trt)) + 
    geom_density(adjust=3) + 
    theme_cowplot() + 
    labs(title=names(slope_reps)[i],x="",y="") +
    theme(legend.position = "none",
          plot.title = element_text(hjust = 0.5))
}
```

### Distribution of slopes for various phenotypes by parental treatment  
```{r,fig.width=10,fig.height=10}
plot_grid(plotlist = plot_slope_list2) 
```


## Option 3 - Regression based on individual samples  
```{r}
pheno_ind <- ind$pheno
pheno_ind <-  pheno_ind[!is.na(pheno_ind$GrowthPerDay),]
coln <- c("JarID","JarpHNBS","parentalComb")
jrp_red <- subset(jrp,select=coln)
phenoID <- left_join(pheno_ind,jrp_red,by="JarID")
phenoID <-  phenoID[!is.na(phenoID$parentalComb),]
y <- c("GrowthPerDay")
x <- "JarpHNBS"
slope_mat <- matrix(ncol=length(y),nrow=length(unique(phenoID$parentalComb)))
colnames(slope_mat) <- y
intercept_mat <- matrix(ncol=length(y),nrow=length(unique(phenoID$parentalComb)))
colnames(intercept_mat) <- y
for( i in 1:length(unique(phenoID$parentalComb))){
  for(j in 1:length(y)){
    temp <- phenoID[phenoID$parentalComb == unique(phenoID$parentalComb)[i],]
    lm_out <- lm(temp[,y[j]]~temp[,x])
    lm_sum <- summary(lm_out)
    intercept_mat[i,j] <- lm_sum$coefficients[1,1]
    slope_mat[i,j] <- lm_sum$coefficients[2,1]
  }
}

ind_growth <- data.frame(parentalComb=unique(phenoID$parentalComb))
ind_growth <- left_join(ind_growth,jrp_pcombo,by="parentalComb")
ind_intercept <- data.frame(ind_growth,intercept_mat)
ind_slope <- data.frame(ind_growth,slope_mat)

ind <- list(intercept=ind_intercept,slope=ind_slope)
```

```{r}
plot_slopes <- list()
plot_intercepts <- list()
for( i in 1:length(y)){
  temp <- data.frame(Parent_Treatment=as.factor(ind$slope$ParentTrt),Slope=ind$slope[,i+2])
  out <- ggplot(temp,aes(Slope,group=Parent_Treatment,colour=Parent_Treatment)) + 
    geom_density() +
    theme_cowplot() +
    labs(title=y[i],x="",y="") +
    theme(legend.position = "none",
          plot.title = element_text(hjust = 0.5))
  plot_slopes[[i]] <- out
  temp <- data.frame(Parent_Treatment=as.factor(ind$intercept$ParentTrt),Intercept=ind$intercept[,i+2])
  out <- ggplot(temp,aes(Intercept,group=Parent_Treatment,colour=Parent_Treatment)) + 
    geom_density() +
    theme_cowplot() +
    labs(title=y[i],x="",y="") +
    theme(legend.position = "none",
          plot.title = element_text(hjust = 0.5))
  plot_intercepts[[i]] <- out
    
}
```

### Distribution of slopes for various phenotypes by parental treatment  
```{r echo=F,fig.width=10,fig.height=10}
plot_grid(plotlist=plot_slopes)
  ```

