---
title: "Estimating Heritability Example - Larval Growth"
output: 
  github_document:
    toc: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include=FALSE}
library(MCMCglmm)
library(ggplot2)
library(ggpubr)
library(cowplot)
library(dplyr)
library(knitr)
```

# Overview 

Here I look at estimating heritability of larvae growth using a bayesian mixed modelling approach implemented in MCMCglmm. I specifically look at the impact of an uninformative prior (inverse-gamma) vs. a slightly more informative prior (Chi2), as well as the effect of including dam, sire, and environment into the model.

After some initial testing I determined that 1 million iterations (`nitt`), 10,000 generation burnin (`burnin=10000`), and 200 generation thinning steps (`thin=200`) were mostly sufficient for reducing autocorrelation (auto correlation < 0.1) and generating a sufficient effective population size (Ne > 2000) for all models.

### Data
```{r}
jr <- readRDS("~/Github/2018OAExp_larvae/input_files/JarHeritabilityData.RDS")
ind <- readRDS("~/Github/2018OAExp_larvae/input_files/IndHeritabilityData.RDS")

ped1 <- jr$ped
jrp <- jr$pheno

# Control Parents - Control Offspring 
jr_cc <- subset(jr$pheno, jr$pheno$JarTrt == "Control" & jr$pheno$ParentTrt == 400)
dam_cc <- data.frame(id=unique(jr_cc$damID),dam=NA,sire=NA)
sire_cc <- data.frame(id=unique(jr_cc$sireID),dam=NA,sire=NA)
ped_cc <- rbind(dam_cc,sire_cc,data.frame(id=jr_cc$animal,dam=jr_cc$damID,sire=jr_cc$sireID))
```

# Estimating Heritability  

###Assumptions of the animal model
1) The Y trait is normally distributed (this assumption can be drop using generalized models)
2) The breeding values ai are normally distributed and correlated among related individuals.
3) The function of the pedigree is to structure the correlation between individuals by taking into account their kinship.
4) The residuals ei are normally distributed and uncorrelated. They also are independent from the breeding values (e.g. no environment-genotype interaction).

Things to remember 
* Scale the phenotypes otherwise estimate of the priors can be errorneous

**Scale Growth**
```{r}
jr_cc$GrowthScale <- scale(jr_cc$Growth)
hist(jr_cc$GrowthScale,main="Scaled Growth",xlab="Scaled Growth")
```
y-variable is mostly normal and is now scaled about 0.

## Priors

### Inverse gamme prior
Classic uninformative prior used in animal model. This prior doesn't inform posterior distribution estimates, meaning estimates shouls be largely determined by the data (usefull when we do not have strong a priori expectations).
```{r}
prior1 <- list(R = list(V=1, nu=0.002), G = list(G1 = list(V=1, nu=0.002)))
```

### Chi square prior 
Can be used when variance in data is small, slightly more informative than inverse gamma prior.
```{r}
prior2<-list(R=list(V=1,nu=1),
                G=list(G1=list(V=1,nu=1,alpha.mu=0,alpha.V=1000)))
```

**Chi-squared with two random effects with identical priors**
```{r}
prior2.2 <- list(R = list(V=1, nu=0.002), G = list(G1 = list(V=1,nu=1,alpha.mu=0,alpha.V=1000),
                                                 G2 = list(V=1,nu=1,alpha.mu=0,alpha.V=1000)))
```
**Chi-squared with three random effects with identical priors**
```{r}
prior2.3<-list(R=list(V=1,nu=1),
             G=list(G1=list(V=1,nu=1,alpha.mu=0,alpha.V=1000),
                    G2=list(V=1,nu=1,alpha.mu=0,alpha.V=1000),
                    G3=list(V=1,nu=1,alpha.mu=0,alpha.V=1000)))
```

# Model 1 - Simple model no added fixed or random effects and uninformative prior  
Estimate heritability of larvae growth (jar-level) in absence of treatment (i.e. control-control).
```{r}
# Simple model with no fixed effects or additional random effects
model_mcmc  <- MCMCglmm(GrowthScale ~ 1,
                        random = ~ animal,
                        family = "gaussian",
                        prior = prior1,
                        pedigree = ped_cc,
                        data = jr_cc,
                        nitt = 1000000,
                        burnin = 10000,
                        thin = 200,
                        verbose = F)
```

```{r echo=F}
## Calculate heritability for heritability from variance components
h2_mcmc_object  <- model_mcmc$VCV[, "animal"] /
  (model_mcmc$VCV[, "animal"] + model_mcmc$VCV[, "units"])

## Summary of model
(m1_sum <- summary(model_mcmc))

## Summarise results from that posterior
h2_m1_mcmc  <- data.frame(method = "Model 1",
                          prior="Inverse-Gamma",
                          fixed_effects=paste0(c(m1_sum$fixed.formula)),
                          random_effects=paste0(c(m1_sum$random.formula)),
                          DIC=m1_sum$DIC,
                          H2_mean = mean(h2_mcmc_object),
                          H2_lower = quantile(h2_mcmc_object, 0.025),
                          H2_upper = quantile(h2_mcmc_object, 0.975),
                          stringsAsFactors = FALSE,
                          row.names = NULL)
```

## Model 1 - Diagnostics
```{r echo=F}
# Traces for fixed effects
plot(model_mcmc$Sol)
# Traces for random effects
plot(model_mcmc$VCV)
## Auto-correlation among lag steps
autocorr.diag(model_mcmc$Sol)
autocorr.diag(model_mcmc$VCV)

## Effective sample size
effectiveSize(model_mcmc$Sol)
effectiveSize(model_mcmc$VCV)

## Test of model convergence
heidel.diag(model_mcmc$VCV)
```

# Model 2 - Changing to slightly more informative prior  
```{r}
# Simple model with no fixed effects or additional random effects
model2_mcmc  <- MCMCglmm(GrowthScale ~ 1,
                        random = ~ animal,
                        family = "gaussian",
                        prior = prior2,
                        pedigree = ped_cc,
                        data = jr_cc,
                        nitt = 1000000,
                        burnin = 10000,
                        thin = 200,
                        verbose = F)
```

```{r echo=F}
## Calculate heritability for heritability from variance components
h2_mcmc_object  <- model2_mcmc$VCV[, "animal"] /
  (model2_mcmc$VCV[, "animal"] + model2_mcmc$VCV[, "units"])
## Summary of model
(m2_sum <- summary(model2_mcmc))

## Summarise results from that posterior
h2_m2_mcmc  <- data.frame(method = "Model 2",
                          prior="Chi-squared",
                          fixed_effects=paste0(c(m2_sum$fixed.formula)),
                          random_effects=paste0(c(m2_sum$random.formula)),
                          DIC=m2_sum$DIC,
                          H2_mean = mean(h2_mcmc_object),
                          H2_lower = quantile(h2_mcmc_object, 0.025),
                          H2_upper = quantile(h2_mcmc_object, 0.975),
                          stringsAsFactors = FALSE,
                          row.names = NULL)

```

## Model 2 - Diagnostics  
```{r echo=F}
# Traces for fixed effects
plot(model2_mcmc$Sol)
# Traces for random effects
plot(model2_mcmc$VCV)

## Auto-correlation among lag steps
autocorr.diag(model2_mcmc$Sol)
autocorr.diag(model2_mcmc$VCV)

## Effective sample size
effectiveSize(model2_mcmc$Sol)
effectiveSize(model2_mcmc$VCV)

## Test of model convergence
heidel.diag(model2_mcmc$VCV)
```

# Model 3 - Adding fixed effects - dam and sire 
```{r}
model3_mcmc  <- MCMCglmm(GrowthScale ~ 1 + damID + sireID,
                         random = ~ animal,
                         family = "gaussian",
                         prior = prior2,
                         pedigree = ped_cc,
                         data = jr_cc,
                         nitt = 1000000,
                         burnin = 10000,
                         thin = 200,
                         verbose = F)
```

```{r echo=F}
## Calculate heritability for heritability from variance components
h2_mcmc_object  <- model3_mcmc$VCV[, "animal"] /
  (model3_mcmc$VCV[, "animal"] + model3_mcmc$VCV[, "units"])
# Summary
(m3_sum <- summary(model3_mcmc))
## Summarise results from that posterior
h2_m3_mcmc  <- data.frame(method = "Model 3",
                          prior="Chi-squared",
                          fixed_effects=paste0(c(m3_sum$fixed.formula)),
                          random_effects=paste0(c(m3_sum$random.formula)),
                          DIC=m3_sum$DIC,
                          H2_mean = mean(h2_mcmc_object),
                          H2_lower = quantile(h2_mcmc_object, 0.025),
                          H2_upper = quantile(h2_mcmc_object, 0.975),
                          stringsAsFactors = FALSE,
                          row.names = NULL)
```

### Model 3 - Diagnostics
```{r echo=F}
plot(model3_mcmc$Sol)
plot(model3_mcmc$VCV)

effectiveSize(model3_mcmc$Sol)
effectiveSize(model3_mcmc$VCV)

heidel.diag(model3_mcmc$VCV)


```

# Model 4 - Including dam and sire as random effects 
```{r}
model4_mcmc  <- MCMCglmm(GrowthScale ~ 1,
                         random = ~ animal + damID + sireID,
                         family = "gaussian",
                         prior = prior2.3,
                         pedigree = ped_cc,
                         data = jr_cc,
                         nitt = 1000000,
                         burnin = 10000,
                         thin = 200,
                         verbose = F)
```

```{r echo=F}
h2_mcmc_object  <- model4_mcmc$VCV[, "animal"] /
  (model4_mcmc$VCV[, "animal"] +  + model4_mcmc$VCV[, "damID"] + model4_mcmc$VCV[, "sireID"] + model4_mcmc$VCV[, "units"])
# Summary
(m4_sum <- summary(model4_mcmc))

h2_m4_mcmc  <- data.frame(method = "Model 4",
                          prior="Chi-squared",
                          fixed_effects=paste0(c(m4_sum$fixed.formula)),
                          random_effects=paste0(c(m4_sum$random.formula)),
                          DIC=m4_sum$DIC,
                          H2_mean = mean(h2_mcmc_object),
                          H2_lower = quantile(h2_mcmc_object, 0.025),
                          H2_upper = quantile(h2_mcmc_object, 0.975),
                          stringsAsFactors = FALSE,
                          row.names = NULL)


```

## Model 4 - Diagnostics
```{r echo=F}
plot(model4_mcmc$Sol)
plot(model4_mcmc$VCV)

effectiveSize(model4_mcmc$Sol)
effectiveSize(model4_mcmc$VCV)

autocorr.diag(model4_mcmc$Sol)
autocorr.diag(model4_mcmc$VCV)

heidel.diag(model4_mcmc$VCV)
```

# Model 5 - Including Egg Diameter and Jar Chemistry as fixed effects 
```{r}
model5_mcmc  <- MCMCglmm(GrowthScale ~ 1 + EggDiamum + JarpCO2_SW,
                         random = ~ animal,
                         family = "gaussian",
                         prior = prior2,
                         pedigree = ped_cc,
                         data = jr_cc,
                         nitt = 1000000,
                         burnin = 10000,
                         thin = 200,
                         verbose = F)
```

```{r echo=F}
## Calculate heritability for heritability from variance components
h2_mcmc_object  <- model5_mcmc$VCV[, "animal"] /
  (model5_mcmc$VCV[, "animal"] + model5_mcmc$VCV[, "units"])
# Summary
(m5_sum <- summary(model5_mcmc))

## Summarise results from that posterior
h2_m5_mcmc  <- data.frame(method = "Model 5",
                          prior="Chi-squared",
                          fixed_effects=paste0(c(m5_sum$fixed.formula)),
                          random_effects=paste0(c(m5_sum$random.formula)),
                          DIC=m5_sum$DIC,
                          H2_mean = mean(h2_mcmc_object),
                          H2_lower = quantile(h2_mcmc_object, 0.025),
                          H2_upper = quantile(h2_mcmc_object, 0.975),
                          stringsAsFactors = FALSE,
                          row.names = NULL)
```

## Model 5 - Diagnostics
```{r echo=F}
plot(model5_mcmc$Sol)
plot(model5_mcmc$VCV)

effectiveSize(model5_mcmc$Sol)
effectiveSize(model5_mcmc$VCV)

autocorr.diag(model5_mcmc$Sol)
autocorr.diag(model5_mcmc$VCV)

heidel.diag(model5_mcmc$VCV)
```

# Model 6 -  Includes eggdiam and pCO2 as fixed effects and damID as random effect
```{r}
model6_mcmc  <- MCMCglmm(GrowthScale ~ 1 + EggDiamum + JarpCO2_SW,
                         random = ~ animal + damID,
                         family = "gaussian",
                         prior = prior2.2,
                         pedigree = ped_cc,
                         data = jr_cc,
                         nitt = 1000000,
                         burnin = 10000,
                         thin = 200,
                         verbose = F)
```

```{r echo=F}
h2_mcmc_object  <- model6_mcmc$VCV[, "animal"] /
  (model6_mcmc$VCV[, "animal"] + model6_mcmc$VCV[, "damID"] + model6_mcmc$VCV[, "units"])
# Model Summary
(m6_sum <- summary(model6_mcmc))

## Summarise results from that posterior
h2_m6_mcmc  <- data.frame(method = "Model 6",
                          prior="Chi-squared",
                          fixed_effects=paste0(c(m6_sum$fixed.formula)),
                          random_effects=paste0(c(m6_sum$random.formula)),
                          DIC=m6_sum$DIC,
                          H2_mean = mean(h2_mcmc_object),
                          H2_lower = quantile(h2_mcmc_object, 0.025),
                          H2_upper = quantile(h2_mcmc_object, 0.975),
                          stringsAsFactors = FALSE,
                          row.names = NULL)
```

## Model 6 - Diagnostics
```{r echo=F}
plot(model6_mcmc$Sol)
plot(model6_mcmc$VCV)

autocorr.diag(model6_mcmc$Sol)
autocorr.diag(model6_mcmc$VCV)

heidel.diag(model6_mcmc$VCV)
```

# Model 7 -  Includes damID as random effect
```{r}
model7_mcmc  <- MCMCglmm(GrowthScale ~ 1,
                         random = ~ animal + damID,
                         family = "gaussian",
                         prior = prior2.2,
                         pedigree = ped_cc,
                         data = jr_cc,
                         nitt = 1000000,
                         burnin = 10000,
                         thin = 200,
                         verbose = F)
```

```{r echo=F}
h2_mcmc_object  <- model7_mcmc$VCV[, "animal"] /
  (model7_mcmc$VCV[, "animal"] + model7_mcmc$VCV[, "damID"] + model7_mcmc$VCV[, "units"])
# Model Summary
(m7_sum <- summary(model7_mcmc))

## Summarise results from that posterior
h2_m7_mcmc  <- data.frame(method = "Model 7",
                          prior="Chi-squared",
                          fixed_effects=paste0(c(m7_sum$fixed.formula)),
                          random_effects=paste0(c(m7_sum$random.formula)),
                          DIC=m7_sum$DIC,
                          H2_mean = mean(h2_mcmc_object),
                          H2_lower = quantile(h2_mcmc_object, 0.025),
                          H2_upper = quantile(h2_mcmc_object, 0.975),
                          stringsAsFactors = FALSE,
                          row.names = NULL)
```

## Model 7 - Diagnostics
```{r echo=F}
plot(model7_mcmc$Sol)
plot(model7_mcmc$VCV)

autocorr.diag(model7_mcmc$Sol)
autocorr.diag(model7_mcmc$VCV)

heidel.diag(model7_mcmc$VCV)
```

# Summary of heritability by model  
```{r echo=F}
summary_table <- rbind(h2_m1_mcmc,h2_m2_mcmc,h2_m3_mcmc,
      h2_m4_mcmc,h2_m5_mcmc,h2_m6_mcmc,h2_m7_mcmc)
kable(summary_table)
```